On hyperparameter tuning through Lipschitz global optimization

### 2024-07-26 2024/ITMO paper/sn-article.tex
before final check

An essential feature of the iOpt framework is the use of Lipschitz optimisation methods for hyperparameter tuning. These methods are deterministic, provide fast convergence, and can be adapted to various types of special problem formulations. For example, such methods allow generalisation for working with noisy functions  \citep{Sergeyev2020}. In this paper, these methods have been adapted to solve HPO problems where some of the parameters are often discrete and/or categorical.

@@

Let us now consider the mapping
\[
Y(x)=y(x-E(x)), \; x\in[0,S],
\]
which maps any point of the interval $[0,S]$ to the region $D$ (the notation $E(x)$ corresponds to the integer part of the number $x$) and define the function
\begin{equation}\label{f}
f(x) = \varphi(Y(x),u_{E(x)+1}), x\in[0,S],
\end{equation}
which has, generally speaking, discontinuities at integer points $x_i = i, 1 \leq i \leq S-1$.
Therefore, the values $z_i = f(x_i)$ at these points will be considered undefined and will not be used in the algorithm.
%, а значения индекса -- равным 0, т.е. $\nu(x_i) = 0$.

@@

The iOpt framework can be used to develop specialized decision support systems for solving the tasks of selecting optimal parameters for complex models or methods.
\textcolor[rgb]{0,0,1}{In particular, it is focused on solving the problems of tuning hyperparameters of machine learning methods. In such problems there is a lack of analytical description of the objective function (``black box'' type problems), and the function itself can be multi-extremal.}

@@

\begin{algorithm}
\caption{Global search algorithm}\label{alg_GSA}
\begin{algorithmic}[1]
\Require $\varphi(y), D, r, \epsilon, K_{max}$
\Ensure $y_{min}$
\State $k \Leftarrow 2$, $t \Leftarrow 2$
\State $v_1 \Leftarrow (x_1 \Leftarrow 0, y_1 \Leftarrow y(x_1), z_1 \Leftarrow \varphi(y_1))$
\State $v_2 \Leftarrow (x_2 \Leftarrow 1, y_2=y(x_2), z_2 \Leftarrow \varphi(y_2))$
\State $\Omega_k \Leftarrow \left\{ v_1, v_2 \right\}$
\While{ $\Delta_t \geq \epsilon$ and $k \leq K_{max}$}
    \State compute the estimate of the H\"older constant $\mu$ by (\ref{mu})
    \For{$i=0$ to $k$}
        \State compute the characteristic $R(i)$ by (\ref{R})
    \EndFor
    \State determine the number $t$, for which $R(t) = \max \left\{ R(i), 1 \leq i \leq k \right\}$
    \State compute the new trial point $x^{k+1}$ by (\ref{xk1})
    \State perform a new trial at the point $x^{k+1}$
    \State $\Omega_k \Leftarrow \Omega_k \cup (x^{k+1}, y^{k+1}, z^{k+1})$
    \State $k \Leftarrow k + 1$
\EndWhile
\State $l \Leftarrow \arg \min \left\{ z_i, 1 < i \leq k \right\}$
\State $y_{min} \Leftarrow y_l$
\State \Return {$y_{min}$}
\end{algorithmic}
\end{algorithm}

@@

\begin{enumerate}[1.]
\item At the initialization stage, the values $(x^i = i, \nu^i=0, z^i)$, $ i, 0\leq i \leq S$, i.e., the boundary points of the intervals $(x_{i-1},x_i), 1<i\leq S$, are entered into the set $\Omega_k$ containing the search  information. 
After that, trials are carried out at arbitrary internal points (e.g., midpoints) of the intervals $(x_{i-1},x_i), 1<i\leq S$; the trial results are also added to the search information. Thus, after initialization, the set $\Omega_k$ will contain $k=2S+1$ elements.

@@

%Новый текст на русском
The convergence conditions of the Global Search Algorithm for problems with Categorical Parameters (GSA-CP) can be formulated as the following statement.
\begin{proposition}
Let the functions $ \varphi(y,u_s), y \in D, s = 1,...,S,$ from (\ref{problem_is}) satisfy the Lipschitz condition with respective constants $L_s, s = 1,...,S$, and $L = \max\{L_s\}$.
Also let the point $\overline{x}$ be a limit point of the sequence $\left\{x^k\right\} $, obtained by minimizing the function $f(x)$ from (\ref{f}) using the GSA-CP algorithm. Then:
\begin{enumerate}
	\item 
	if at the same time with $\overline{x}$ there exists another limit point $x'$ of the sequence $\left\{x^k\right\}$, then $f(\overline{x}) = f(x')$;
	\item
	$z^k=f(x^k) \geq f(\overline{x})$ for any  $k \geq 0$;
	\item
	If at some search step the value $\mu$ from (\ref{mu}) satisfies the condition
\begin{equation}\label{217}
  r \mu > 2^{3-1/N} L \sqrt{N+3},
\end{equation}
then $\overline{x}$ is a global minimizer of the function $f(x), x\in[0,S]$ and any global minimizer $x^\ast$ is also a limit point of the sequence $\left\{x^k\right\}$.
\end{enumerate}
\end{proposition}
\begin{proof}
This statement is proved in a similar way to the theorem on the convergence of the index method of global optimisation from \citet{Strongin2000}, taking into account the relationship (\ref{f}) of the functions $f(x)$ and $\varphi(y,u_s)$, as well as the rules (\ref{index}), specifying the indices of the trial points.
\end{proof}

>>>>

An essential feature of the iOpt framework is the use of Lipschitz optimization methods for hyperparameter tuning. These methods are deterministic, provide fast convergence, and can be adapted to various types of special problem formulations. For example, such methods allow generalisation for working with noisy functions  \citep{Sergeyev2020}. In this paper, these methods have been adapted to solve HPO problems where some of the parameters are often discrete and/or categorical.

@@

Let us now consider the mapping
\[
Y(x)=y(x-E(x)), \; x\in[0,S],
\]
which maps any point of the interval $[0,S]$ to the region $D$ (the notation $E(x)$ corresponds to the integer part of the number $x$) and define the function
\begin{equation}\label{f}
f(x) = \varphi(Y(x),u_{E(x)+1}), x\in[0,S],
\end{equation}
which has, generally speaking, discontinuities at integer points $x_i = i, 1 \leq i \leq S-1$.
Therefore, the values $z_i = f(x_i)$ at these points will be considered undefined and will not be used in the algorithm.
\textcolor[rgb]{0,0,1}{For the same treatment of boundary points of the interval $[0,S]$ we will also consider the values $f(0)$ and $f(S)$ to be undefined. The function values at boundary points will not be used in the algorithm.}

@@

The iOpt framework can be used to develop specialized decision support systems for solving the tasks of selecting optimal parameters for complex models or methods.
\textcolor[rgb]{0,0,1}{In particular, it is focused on solving the problems of tuning hyperparameters of machine learning methods. In such problems there is a lack of analytical description of the objective function and the function itself can be multi-extremal.}

@@

\begin{algorithm}
\caption{Global search algorithm}\label{alg_GSA}
\begin{algorithmic}[1]
\Require $\varphi(y), D, r, \epsilon, K_{max}$
\Ensure $y_{min}$
\State $k \Leftarrow 2$, $t \Leftarrow 2$
\State $v_1 \Leftarrow (x_1 \Leftarrow 0, y_1 \Leftarrow y(x_1), z_1 \Leftarrow \varphi(y_1))$
\State $v_2 \Leftarrow (x_2 \Leftarrow 1, y_2=y(x_2), z_2 \Leftarrow \varphi(y_2))$
\State $\Omega_k \Leftarrow \left\{ v_1, v_2 \right\}$
\While{ $\Delta_t \geq \epsilon$ and $k \leq K_{max}$}
    \State compute the estimate of the H\"older constant $\mu$ by (\ref{mu})
    \For{$i=2$ to $k$}
        \State compute the characteristic $R(i)$ by (\ref{R})
    \EndFor
    \State determine the number $t$, for which $R(t) = \max \left\{ R(i), 1 < i \leq k \right\}$
    \State compute the new trial point $x^{k+1}$ by (\ref{xk1})
    \State perform a new trial at the point $x^{k+1}$
    \State $\Omega_k \Leftarrow \Omega_k \cup (x^{k+1}, y^{k+1}, z^{k+1})$
    \State $k \Leftarrow k + 1$
\EndWhile
\State $l \Leftarrow \arg \min \left\{ z_i, 1 < i \leq k \right\}$
\State $y_{min} \Leftarrow y_l$
\State \Return {$y_{min}$}
\end{algorithmic}
\end{algorithm}

@@

\begin{enumerate}[1.]
\item At the initialization stage, the values $(x^i = i, \nu^i=0, z^i)$, $ 0\leq i \leq S$, i.e., the boundary points of the intervals $(x_{i-1},x_i), 1<i\leq S$, are entered into the set $\Omega_k$ containing the search  information.
\textcolor[rgb]{0,0,1}{Recall that the values $z^i$ at integer points $x^i=i, 0\leq i \leq S,$ are supposed to be undefined and are not used later in the algorithm.}
After that, trials are carried out at arbitrary internal points (e.g., midpoints) of the intervals $(x_{i-1},x_i), 1<i\leq S$; the trial results are also added to the search information. Thus, after initialization, the set $\Omega_k$ will contain $k=2S+1$ elements.

@@

%Новый текст на русском
The convergence conditions of the Global Search Algorithm for problems with Categorical Parameters (GSA-CP) can be formulated as the following statement.
\begin{proposition}
Let the functions $ \varphi(y,u_s), y \in D, s = 1,...,S,$ from (\ref{problem_is}) satisfy the Lipschitz condition with respective constants $L_s, s = 1,...,S$, and $L = \max\{L_s\}$.
Also let the point $\overline{x}$ be a limit point of the sequence $\left\{x^k\right\} $, obtained by minimizing the function $f(x)$ from (\ref{f}) using the GSA-CP algorithm. Then:
\begin{enumerate}
	\item 
	if at the same time with $\overline{x}$ there exists another limit point $x'$ of the sequence $\left\{x^k\right\}$, then $f(\overline{x}) = f(x')$;
	\item
	$z^k=f(x^k) \geq f(\overline{x})$ for any  $k \geq 0$;
	\item
	If at some search step the value $\mu$ from (\ref{mu}) satisfies the condition
\begin{equation}\label{217}
  r \mu > 2^{3-1/N} L \sqrt{N+3},
\end{equation}
then $\overline{x}$ is a global minimizer of the function $f(x), x\in[0,S]$ and any global minimizer $x^\ast$ is also a limit point of the sequence $\left\{x^k\right\}$.
\end{enumerate}
\end{proposition}
\begin{proof}
This statement is proved in a similar way to the theorem on the convergence of the index method of global optimization from \citet{Strongin2000}, taking into account the relationship (\ref{f}) of the functions $f(x)$ and $\varphi(y,u_s)$, as well as the rules (\ref{index}), specifying the indices of the trial points.
\end{proof}
